Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-08 _tex/paper2.tex
Built site for gh-pages

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \((X_{I})\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \((T_{I})\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) columns (variables), and both \(e_{I}\) and \(T_{I}\) are
latent vectors of length \(n_{I}\). Additionally, this branch shows how
\(T_{I}\), along with \emph{relevant} \footnote{refer to footnote 2.}
text-related variables \((X_{IA})\) (e.g., text length), and texts'
idiosyncratic errors \(e_{IA}\) causally influence the texts'
writing-quality trait \((T_{IA})\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
are latent vectors of length \(n_{I} \cdot n_{A}\).

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
columns (variables), and both \(e_{J}\) and \(B_{J}\) are latent vectors
of length \(n_{J}\). Furthermore, the branch shows how \(B_{J}\), along
with \emph{relevant} \footnote{refer to footnote 2.} judgment-related
variables \((Z_{JK})\) (e.g., the number of judgments a judge makes),
and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
biases associated with each text \(B_{JK}\), the second primary latent
variable of interest. The error \(e_{JK}\) captures variations in
judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent vectors
of length \(n_{J} \cdot n_{K}\).

@@

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. In
Section~\ref{sec-theory-theoretical_SC1}, it explicitly assumes
researchers work with a data sample consisting of a limited number of
repeated judgments \((n^{s}_{K})\) from a sample of judges
\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
sample of students \((n^{s}_{I})\), all drawn from the conceptual
population. Moreover, in Section~\ref{sec-theory-theoretical_SC2}, the
model assumes that judges do not perform \emph{all repeated judgments}
within the data sample. Instead, they conduct a sufficient number of
stimuli comparisons, \(n_{C}\), to ensure an accurate estimation of the
proportion \(P(B>A)\).

@@

Where each element of \(S_{I}\) is a binary value indicating the
presence or absence of corresponding elements in the vector \(I\), as in
equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\), based
on the vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and
\(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements, respectively.
\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is sampled} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
}\end{equation}

Similarly, each element of \(S_{IA}\) is a binary value indicating the
presence or absence of data rows in the matrices \(IA\), as defined in
equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\), based
on the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
\(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
elements, respectively. \begin{equation}\phantomsection\label{eq-mat23}{
ia_{(i,a)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are sampled} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
}\end{equation}

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge. They also omit the second text from each
individual and the third repeated judgment from each judge. Given
\(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
vectors take the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

@@

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses this vector, indicating that it is a conditioned variable.
Here, \emph{conditioning} means that researchers restrict their focus to
the elements of \(O^{cp}_{R}\) that satisfy \(s_{(h,i,a,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
sample equivalent for the \(R\) matrix, that is, this vector selects
\emph{all repeated judgments a subset of judges makes for the subset of
stimuli produced by the sampled individuals}. This process generates the
sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
outcome \(O^{cp}_{R}\).

@@

However, because CJ experiments rarely use the exhaustive pairing of
sampled judges, stimuli, and individuals--due to concerns about the
practical feasibility of the comparison task \citep{Boonen_et_al_2020}--
a realistic scenario must also account for judges comparing only
specific stimuli from certain individuals.

@@

Where \(C\) contains \(n\) elements corresponding to the number of rows
in the \(R\) matrix, with each element \(c_{(h,a,i,b,j,k)}\) being a
binary value indicating the presence or absence of data rows in the
\(R\) similar to equation (\ref{eq-mat26}). This definition assumes that
the subset of judges \emph{does not} perform all repeated judgments for
the subset of stimuli produced by the sampled individuals. Instead, they
conduct a ``sufficient number of comparisons'' \(n_{C}\) to ensure the
accurate estimation of the proportion \(P(B>A)\) for each pair of
stimuli \citep[pp.~267]{Thurstone_1927b}.

DAG~\ref{fig-cj14_dag} also incorporates the comparison mechanism \(C\)
into the conceptual-population model. The DAG represents the
sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
that researchers cannot directly access this outcome due to the
comparison mechanism. Moreover, it represents the \emph{comparison
mechanism} \(C\) as a causal factor influencing the final comparative
outcome \(O_{R}\). Equivalent to the sampling mechanism, a square
encloses this vector, indicating that \(C\) is a conditioned variable
that determines \emph{which repeated judgments the subset of judges
makes for the subset of stimuli produced by the sampled individuals}.
This process ultimately derives the final comparative outcome \(O_{R}\)
from the sample-comparison outcome \(O^{sc}_{R}\).

@@

Finally, without loss of generality, researchers can reformulate the
sample-comparison model in Figure~\ref{fig-cj14} into the equivalent
form shown in Figure~\ref{fig-cj15}. This reformulation rests on two
principles. First, because the sampling and comparison mechanism vectors
\(S\) and \(C\) are binary, researchers can integrate them into the
model using element-wise products. This approach eliminates the need to
define the structural equations \(f_{S}\) and \(f_{C}\). Second, since
only the final comparative outcome \(O_{R}\) is observed, it is standard
to assume that the distribution of the conceptual-population outcome
\(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
applies to this outcome \(O_{R}\). This assumption removes the necessity
of describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).

@@

Using SCM~\ref{fig-cj15_scm} we finally define the

@@

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
  D_{R} & = \left( T_{IA}[R_{1},R_{2}] - T_{IA}[R_{3},R_{4}] \right) + B_{JK}[R_{5},R_{6}] \\
  T_{IA} & = T_{I}[IA_{1}] + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J}[JK_{1}] + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
\]

@@

\subcaption{\label{fig-cj16_scm}SCM}

@@

@@

A notable implication of the assumptions underlying the probabilistic
encoding is that any conceptual model described by an SCM and DAG can
represent the joint distribution of variables more efficiently
\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
product of conditional probability distributions (CPDs) of the type
\(P(child \:|\:parents)\). This property is formally known as the
\emph{Bayesian Network factorization} (BNF,
Equation~\ref{eq-net_factor})
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this expression, \(pa(X_{i})\) denotes the set of variables that are
the parents of \(X_{i}\).

\begin{equation}\phantomsection\label{eq-net_factor}{
\begin{aligned}
P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:pa(X_{i}) ) & (\small{\text{by BNF}})
\end{aligned}
}\end{equation}

>>>>

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \((X_{I})\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \((T_{I})\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) columns (variables), and both \(e_{I}\) and \(T_{I}\) are
latent vectors of length \(n_{I}\). Additionally, this branch shows how
\(T_{I}\), along with \emph{relevant} \footnote{refer to footnote 2.}
text-related variables \((X_{IA})\) (e.g., text length), and texts'
idiosyncratic errors \(e_{IA}\) causally influence the texts'
writing-quality trait \((T_{IA})\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
are latent matrices with \(n_{I}\) rows and \(n_{A}\) columns.

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
columns (variables), and both \(e_{J}\) and \(B_{J}\) are latent vectors
of length \(n_{J}\). Furthermore, the branch shows how \(B_{J}\), along
with \emph{relevant} \footnote{refer to footnote 2.} judgment-related
variables \((Z_{JK})\) (e.g., the number of judgments a judge makes),
and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
biases associated with each text \(B_{JK}\), the second primary latent
variable of interest. The error \(e_{JK}\) captures variations in
judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent
matrices with \(n_{J}\) rows and \(n_{K}\) columns

@@

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. In
Section~\ref{sec-theory-theoretical_SC1}, it explicitly assumes
researchers work with a data sample consisting of a limited number of
repeated judgments \((n^{s}_{K})\) from a sample of judges
\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
sample of students \((n^{s}_{I})\), all drawn from the conceptual
population. Furthermore, in Section~\ref{sec-theory-theoretical_SC2},
the model assumes that judges do not perform \emph{all repeated
judgments} within the data sample. Instead, they conduct a sufficient
number of stimuli comparisons, \(n_{C}\), to ensure an accurate
estimation of the proportion \(P(B>A)\), as proposed by
\citet{Thurstone_1927b}.

@@

Where each element of \(S_{I}\) is a binary value indicating the
presence or absence of corresponding elements in the vector \(I\), as in
equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\) using
vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and \(S_{J}\)
contains \(n_{I}\) and \(n_{J}\) elements, respectively.
\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is sampled} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
}\end{equation}

Similarly, each element of \(S_{IA}\) is a binary value indicating the
presence or absence of data rows in the matrices \(IA\), as defined in
equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\) using
the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
\(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
elements, respectively. \begin{equation}\phantomsection\label{eq-mat23}{
ia_{(i,a)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are sampled} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
}\end{equation}

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge, and they also omit the second text from
each individual and the third repeated judgment from each judge. Given
\(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
vectors take the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

@@

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses this vector, indicating that it is a conditioned variable.
Here, \emph{conditioning} means that researchers restrict their focus to
the elements of \(O^{cp}_{R}\) that satisfy \(s_{(h,i,a,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
vector selecting \emph{all repeated judgments a subset of judges makes
for the subset of stimuli produced by the sampled individuals}. This
process generates the sample-comparison outcome \(O^{sc}_{R}\).

@@

However, because CJ experiments rarely use the exhaustive pairing of
sampled judges, stimuli, and individuals--due to concerns about the
practical feasibility of the comparison task
\citep{Boonen_et_al_2020}--a realistic scenario must also account for
judges comparing only specific stimuli from certain individuals.

@@

Where \(C\) contains \(n\) elements corresponding to the number of rows
in the \(R\) matrix, with each element \(c_{(h,a,i,b,j,k)}\) being a
binary value indicating the presence or absence of data rows in the
\(R\) similar to equation (\ref{eq-mat26}). This definition assumes that
the sampled judges \emph{do not} perform all repeated judgments for the
subset of stimuli produced by the sampled individuals. Instead, they
conduct a ``sufficient number of comparisons'' \(n_{C}\) to ensure the
accurate estimation of the proportion \(P(B>A)\) for each pair of
stimuli \citep[pp.~267]{Thurstone_1927b}.

DAG~\ref{fig-cj14_dag} also incorporates the comparison mechanism \(C\)
into the conceptual-population model. The DAG represents the
sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
that researchers cannot directly access this outcome due to the
comparison mechanism. Moreover, it represents the \emph{comparison
mechanism} \(C\) as a causal factor influencing the final comparative
outcome \(O_{R}\). Equivalent to the sampling mechanism, a square
encloses this vector, indicating that \(C\) is a conditioned variable
that determines \emph{which repeated judgments the subset of judges
makes for the subset of stimuli produced by the sampled individuals}.
This process ultimately derives the final comparative outcome \(O_{R}\).

@@

Finally, without loss of generality, researchers can reformulate the
sample-comparison model in Figure~\ref{fig-cj14} into the equivalent
form shown in Figure~\ref{fig-cj15}. This reformulation rests on two
principles. First, because the sampling and comparison mechanism vectors
\(S\) and \(C\) are binary, researchers can integrate them into the
model using element-wise products. This approach eliminates the need to
define the structural equations \(f_{S}\) and \(f_{C}\). Second, since
only the final comparative outcome \(O_{R}\) is observed, it is standard
to assume that the distribution of the conceptual-population outcome
\(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
applies to this outcome. This assumption removes the necessity of
describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).

@@

Researchers can easily derive a statistical model to address violations
of the equal dispersion assumption in CJ experiments (see
Section~\ref{sec-theory-issue1a}) using only SCM~\ref{fig-cj15_scm}.
This derivation is possible because the structural equations \(F\)
encode probabilistic information \citep{Pearl_et_al_2016}, allowing
researchers to replace them with suitable probabilistic assumptions. As
a first step, given the dichotomous nature of CJ outcomes, the model
\ref{fig-cj16_stat} assumes that \(O_{R}\) follows a Bernoulli
distribution\footnote{The binomial distribution--including its special
  case, the Bernoulli distribution-- ``represents the maximum entropy
  way to count binary events'' \citep[pp.~34]{McElreath_2020}. This
  means that when only two un-ordered outcomes are possible, and the
  expected numbers of each type of event are assumed to be constant,
  then the binomial distribution is the one that is most consistent with
  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}. Next, following the
conventions of Generalized Linear Latent And Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
model links the observed outcome to the latent discriminal difference
vector \(D_{R}\) through an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Notice that the sampling
and comparison mechanisms, \(S\) and \(C\), act as element-wise
multipliers of the discriminal difference (denoted by the symbol
\(\odot\)), incorporating the design structure into the statistical
formulation. Researchers can omit these vectors when only the observed
CJ sample is available.

@@

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
  D_{R} & = \left( T_{IA}[h,a] - T_{IA}[i,b] \right) + B_{JK}[j,k] \\
  T_{IA}[i,b] & = T_{I}[i] + \beta_{XA} X_{IA}[i,] + e_{IA}[i,b] \\
  T_{I}[i] & = \beta_{XI} X_{I}[i,] + e_{I}[i] \\
  B_{JK}[j,k] & = B_{J}[j] + \beta_{ZK} Z_{JK}[j,] + e_{JK}[j,k] \\
  B_{J}[j] & = \beta_{ZJ} Z_{J}[j,] + e_{J}[j] \\
  e_{I}[i] & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J}[j] & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA}[i,b] & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK}[j,k] & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
\]

@@

\subcaption{\label{fig-cj16_scm}SCM~\ref{fig-cj15_scm}}

@@

Furthermore, the statistical model defines the discriminal difference
\(D_{R}\) as the difference between the writing-quality traits of two
distinct texts, \(T_{IA}[h,a]\) and \(T_{IA}[i,b]\), along with the
corresponding repeated judge biases, \(B_{JK}[j,k]\). The square
brackets indicate the indices of the elements within their respective
trait vectors and matrices, as in equation (\ref{eq-mat11}) and
(\ref{eq-mat12}).

@@

A notable implication of the assumptions underlying the probabilistic
encoding is that any conceptual model described by an SCM and DAG can
represent the joint distribution of variables more efficiently
\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
product of conditional probability distributions (CPDs) of the type
\(P(child \:|\:parents)\). This property is formally known as the
\emph{Bayesian Network factorization} (BNF,
Equation~\ref{eq-net_factor})
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this expression, \(pa(X_{i})\) denotes the set of variables that are
the parents of \(X_{i}\).
\begin{equation}\phantomsection\label{eq-net_factor}{
\begin{aligned}
P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:pa(X_{i}) ) & (\small{\text{by BNF}})
\end{aligned}
}\end{equation}
